Material Culture Networks in Tonto Basin

### 2023-01-11 manuscript.Rmd
updated with new references and discussion

---
title: Material Culture Networks in Tonto Basin
output:
  rticles::tf_article: null
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
author:
  - name: Robert J. Bischoff
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type. Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities. This study takes architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data into a multilayer network analysis. This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation. The results demonstrate a strong association between projectile points and spatial distance. Sites containing roomblock architecture are highly central in the ceramic network but have low centrality in the projectile point network. Overall, the ceramic and projectile point networks exhibit significant differences. This indicates that the social networks that created these patterns had different social mechanisms. One potential cause of these differences is gendered spheres of interaction with men exchanging projectile points and women exchanging ceramics.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
always_allow_html: true
---

@@

Network science has many applications for archaeologists. It can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb]. Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh]. Social systems are complex webs of interrelations. There is too much involved to model the material culture of even a technologically simple society. But certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand. Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture. For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large scale social networks. Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region. This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework. Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated. This is an important question because archaeologists often make inferences based solely on ceramics or other types of material culture. Evidence regarding the ways material culture co-vary will help archaeologists direct future research. Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.

Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub]. This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and a recent projectile point analysis [@Bischoff2022-rg]. There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites. In reality, I expected significant variation between these types of material culture. As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance. Furthermore, ceramic and projectile point networks (hereafter point networks) exhibit significant differences. I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.

@@

One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze]. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu]. This migration into the Tonto Basinwas part of a larger migration from north to south and is associated with the origin of the Salado phenomenon. Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd]. The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks. Having high centrality means they have more connections to other nodes (i.e., sites). The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between sites [e.g., @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

@@

Network approaches are best suited where the relationships between entities are of primary interest. In this case, the entities are defined as archaeology sites. The archaeology sites in turn represent a group of individuals who lived or visited the area of the site and left material remains behind. Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity. Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study. His analysis of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin. He found strong connections between many parts of the eastern Tonto Basin. Unfortunately, this study will only attempt a site-level analysis. Watt's study does, however, illustrate how relations can be defined between nodes in a network analysis. He used similarity networks where the points that had similar flaking styles were connected together. Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd]. These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts. This study also uses similarity networks to group sites by similarity in ceramic assemblages and similarity in projectile point forms.

@@

In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity. The architecture discussed is highly visible and representative of historical group membership. The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership. What this means is that links between nodes in the cer The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study. Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity. He identifies certain patterns, but his synthesis and other research [@Carr1995-mq; @Carr1995-sh; @Dietler1998-zo; @Gosselain1998-qd; @Gosselain2000-wf; @Gosselain2016-as; @Hodder1982-no; @Huntley2008-qu; @Lemonnier1986-le; @Lyons2003-yy; @Neuzil2008-zd; @Sassman2001-sw; @Stark1998-mu; @Wiessner1983-ei; @Wiessner1997-su] strongly indicates that the relationship between material culture and identity is culturally relative. Projectile points can in some cases indicate group membership [@Wiessner1983-ei; @Wiessner1997-su], but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww; @Sliva2002-oz]. Styles were difficult to determine for the original researchers [@Rice1994-rk], and I had the same difficulty. In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case. This is indicative of relational identity. There are no environmental reasons to assume differences in projectile points, as each site was in a similar ecozone. Each hunter would have been using the points for the same game and differences in point styles would have been primarily cultural [@Sliva2002-oz].

There is one other aspect to these relations that I believe played an important role. A central way people identify themselves is by gender. Gender roles, like identity, can also vary in complex ways, but for the purposes of this analysis I will use a simplified model. Women made pottery and men made projectile points. This was not always true of course, but this fits the available data and expectations for the Hohokam [@Crown1996-xb; @Harry2010-hj; @VanPool2010-im, p. 253; @Whittlesey2010-ox]. An examination of the ethnographic record for the O'odham people, recognized descendants of the Hohokam, provides additional references that women made pottery [@Bahr2011-or, p.4; @Castetter1935-pq, p. 5-6; @Chona1936-mg, p. 44; @Joseph1949-ga, p. 57]. An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7]. On the other hand, projectile points were almost exclusively buried with males. This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.

@@

```{r sites, echo=FALSE}
tbls = tbls + 1
cap = "Project Sites and Data"
sites %>% 
  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
  # as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>% 
  # theme_article() %>% 
  # set_caption(cap)
  knitr::kable(caption= cap,
               booktabs = T,
               format.args = list(big.mark = ",",
                                  scientific = FALSE)) %>%
  kable_styling(latex_options = "scale_down")
```

The types of similarity used in this study are varied depending on the type of network. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj], but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.

@@

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network). Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results. Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality. In this case the centrality measure is simply the mean of the eigenvector centraliy for each separate layer, as the layers are not interdependent [@Frost2022-lr]. Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers. These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.

The robustness of results were tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value. This provides a baseline that determines how likely it is to obtain a given result by chance.

Certain caveats must be acknowledged. All archaeologists deal with missing data, which can greatly affect the results of analysis. Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik]. Regardless of statistical methods, unaccounted for biases will still produce invalid results [@Bischoff2021-sf]. The most important bias that may affect these findings is chronology. For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq]. Does that invalidate the network analysis? I do not think so. While the networks involve nodes as sites, what the networks are really meant to represent are groups of people. Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation. Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p. 336]. Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location. The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96]. If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on. It is these communities that the network analysis is attempting to capture. Still, evidence indicates that most sites were contemporaneous. For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya]. It is best, though, to think of these interpretations as estimations based on incomplete data.

@@

Figure `r figs` shows the networks with the top five strongest ties between each node. This number of ties represents the fewest ties that connects all nodes in the spatial network. Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network. Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network. Because the network is not complete, one cannot argue that there were not other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location. Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm. The Rock Island area consists of a single site, Bass Point Mound (a platform mound), although it was not heavily excavated. If spatial distance was an important factor in social interaction, then we would expect Bass Point Mound to consistently be a highly central node due to its central location. Table `r tbls` shows the multilayer eigenvector centrality for points and ceramics only. Bass Point Mound has the highest multilayer network centrality with 0.95 (p = `r pF(.95,c("ceramics","points"),df = "sensitivityEigen")`). Figure `r figs + 1` provides the eigenvector data by type of network. Bass Point Mound has low centrality in the architectural network and the networks that do not connect the spatial components (three and four closest connections), which gives it a lower spatial centrality as well. The ceramic and point networks provide strong evidence that Bass Point Mound's spatial location was advantageous for forming connections between the sites in this study.

```{r TontoNetworksMultiNet, fig.cap= 'Graphs showing the networks used in this analsyis. Each network is displayed with the five strongest ties between each node. Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoNetworksMultiNet.png"))
```

```{r meanEigen, echo=FALSE}
tbls = tbls + 1
cap = "Multilayer Eigenvector Centrality for Ceramics and Points"
meanEigenValsSites %>% 
  select(Site,Centrality) %>% 
  arrange(desc(Centrality)) %>% 
  knitr::kable(caption= cap,
               booktabs = T,
               format.args = list(big.mark = ",",
                                  scientific = FALSE))
  # as_hux() %>% 
  # theme_article() %>% 
  # set_caption(cap)

@@

Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research. Chris Caseldine's comments greatly improved this paper. All mistakes are my own. The many individuals who contributed to the Roosevelt Platform Mound Study also deserve recognition.

>>>>

---
title: Material Culture Networks in Tonto Basin
output:
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
  rticles::tf_article: null
author:
  - name: Robert J. Bischoff
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type. Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities. This study uses architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data in a multilayer network analysis. This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation. The results demonstrate a strong association between projectile points and spatial distance. Overall, the ceramic and projectile point networks exhibit significant differences. This indicates that the social networks that created these patterns had different social mechanisms. One potential cause of these differences is gendered spheres of interaction with men producing and exchanging projectile points and women producing and exchanging ceramics.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
always_allow_html: true
---

@@

Network science has many applications for archaeologists, and it can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb]. Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh]. Social systems are complex webs of interrelations. There is too much involved to model the material culture of even a technologically simple society, but certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand. Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture. For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large-scale social networks. Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region. This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework. Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated. This is important because archaeologists often make inferences based solely on ceramics or other types of material culture. Evidence regarding the ways material culture co-vary will help give better context to these studies. Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.

Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub]. This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and data from a recent projectile point analysis [@Bischoff2022-rg]. There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites. In reality, I expected significant variation between these types of material culture. As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance. Additionally, ceramic and projectile point networks (hereafter point networks) exhibit significant differences. I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.

@@

One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze]. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu], although some have attributed the architectural changes to warfare or population aggregation [c.f. @Oliver2001-tu; @Wood2000-ze]. This migration into the Tonto Basin was part of a larger migration from north to south and is associated with the origin of the Salado phenomenon. Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd]. The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks. Having high centrality means they have more connections to other nodes (i.e., sites). The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between communities in the Tonto Basin [e.g., @Sliva2002-oz; @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

@@

Network approaches are best suited where the relationships between entities are of primary interest. In this case, the entities are defined as archaeology sites. The archaeology sites in turn represent a group of individuals who lived or visited the area of the site and left material remains behind. Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity. Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study. His analysis of flake scar patterning identified individual knappers, or at least clusters of people with similar knapping styles, and the distribution of their projectile points around Tonto Basin. He found strong connections between many parts of the eastern Tonto Basin. Unfortunately, this study will only attempt a site-level analysis. Watt's study does, however, illustrate how relations can be defined between nodes in a network analysis. He used similarity networks where the points that had similar flaking styles were connected together. Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd]. These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts. This study also uses similarity networks to group sites by similarity in ceramic assemblages and similarity in projectile point forms.

@@

In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity. The architecture discussed is highly visible and representative of historical group membership. The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership. What this means is that links between nodes in the ceramic network are more likely to indicate belonging to a similar social group. The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study. Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity. He identifies certain patterns, but his synthesis and other research [@Carr1995-mq; @Carr1995-sh; @Dietler1998-zo; @Gosselain1998-qd; @Gosselain2000-wf; @Gosselain2016-as; @Hodder1982-no; @Huntley2008-qu; @Lemonnier1986-le; @Lyons2003-yy; @Neuzil2008-zd; @Sassman2001-sw; @Stark1998-mu; @Wiessner1983-ei; @Wiessner1997-su] strongly indicates that the relationship between material culture and identity is culturally relative. Projectile points can in some cases indicate group membership [@Wiessner1983-ei; @Wiessner1997-su], but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww; @Sliva2002-oz]. Styles were difficult to determine for the original researchers [@Rice1994-rk], and I had the same difficulty. In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case. This is indicative of relational identity. There are no environmental reasons to assume differences in projectile points, as each site was in a similar ecozone. Each hunter would have been using the points for the same game and differences in point styles would have been primarily cultural [@Sliva2002-oz].

There is one other aspect to these relations that I believe played an important role. A central way people identify themselves is by gender. Gender roles, like identity, can also vary in complex ways, but for the purposes of this analysis I will use a simplified model. Women made pottery and men made projectile points. This was not always true of course, but this fits the available data and expectations for the Hohokam [@Crown1996-xb; @Harry2010-hj; @Shackley2005-hn; @VanPool2010-im, p. 253; @Whittlesey2010-ox]. An examination of the ethnographic record for the O'odham people, recognized descendants of the Hohokam, provides additional references that women made pottery [@Bahr2011-or, p.4; @Castetter1935-pq, p. 5-6; @Chona1936-mg, p. 44; @Joseph1949-ga, p. 57]. An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7]. On the other hand, projectile points were almost exclusively buried with males. This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.

Shackley [-@Shackley2005-hn], using projectile point typology data from Hoffman [-@Hoffman1997-hb] as well as obsidian provenience and ceramic data for preclassic Hohokam sites in the Phoenix Basin, found evidence for three distinct projectile point traditions. Shackley and Hoffman believe these may correlate with warrior sodalities, and Shackley argues that the male projectile point exchange systems functioned in different ways than the female ceramic exchange systems. In this case, the projectile points likely correlate with a categorical identity, but this provides an example of how men and women's networks have been argued to vary in Hohokam archaeology.

@@

```{r sites, echo=FALSE}
tbls = tbls + 1
cap = "Project Sites and Data"
sites %>% 
  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
  as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>%
  theme_article() %>%
  set_caption(cap)
  # knitr::kable(caption= cap,
  #              booktabs = T,
  #              format.args = list(big.mark = ",",
  #                                 scientific = FALSE)) %>%
  # kable_styling(latex_options = "scale_down")
```

The types of similarity used in this study are varied depending on the type of network. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj] but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.

@@

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network). Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results. Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality. In this case the centrality measure is simply the mean of the eigenvector centrality for each separate layer, as the layers are not interdependent [@Frost2022-lr]. Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers. These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.

The robustness of results was tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value. This provides a baseline that determines how likely it is to obtain a given result by chance.

Certain caveats must be acknowledged. All archaeologists deal with missing data, which can greatly affect the results of analysis. Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik]. Regardless of statistical methods, unaccounted for biases will still produce invalid network results [@Bischoff2021-sf]. The most important bias that may affect these findings is chronology. For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq]. Does that invalidate the network analysis? I do not think so. While the networks involve nodes as sites, what the networks are really meant to represent are groups of people. Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation. Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p. 336]. Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location. The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96]. If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on. It is these communities that the network analysis is attempting to capture. Still, evidence indicates that most sites were contemporaneous. For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya]. It is best, though, to think of these interpretations as estimations based on incomplete data.

@@

Figure `r figs` shows the networks with the top five strongest ties between each node. This number of ties represents the fewest ties that connects all nodes in the spatial network. Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network. Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network. Because the network is not complete, one cannot argue that there were no other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location. Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm. The Rock Island area consists of a single site, Bass Point Mound (a platform mound), although it was not heavily excavated. If spatial distance was an important factor in social interaction, then we would expect Bass Point Mound to consistently be a highly central node due to its central location. Table `r tbls` shows the multilayer eigenvector centrality for points and ceramics only. Bass Point Mound has the highest multilayer network centrality with 0.95 (p = `r pF(.95,c("ceramics","points"),df = "sensitivityEigen")`). Figure `r figs + 1` provides the eigenvector data by type of network. Bass Point Mound has low centrality in the architectural network and the networks that do not connect the spatial components (three and four closest connections), which gives it a lower spatial centrality as well. The ceramic and point networks provide strong evidence that Bass Point Mound's spatial location was advantageous for forming connections between the sites in this study.

```{r TontoNetworksMultiNet, fig.cap= 'Graphs showing the networks used in this analysis. Each network is displayed with the five strongest ties between each node. Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoNetworksMultiNet.png"))
```

```{r meanEigen, echo=FALSE}
tbls = tbls + 1
cap = "Multilayer Eigenvector Centrality for Ceramics and Points"
meanEigenValsSites %>% 
  select(Site,Centrality) %>% 
  arrange(desc(Centrality)) %>% 
  # knitr::kable(caption= cap,
  #              booktabs = T,
  #              format.args = list(big.mark = ",",
  #                                 scientific = FALSE))
  as_hux() %>%
  theme_article() %>%
  set_caption(cap)

@@

Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research. Chris Caseldine's comments greatly improved this paper. All mistakes are my own. The many individuals who conducted the Roosevelt Platform Mound Study deserve recognition for contributing the data.
